---
title: "Results for VDBA Scaling Analysis"
author: "O.W."
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.width = 20,
  fig.height = 10
)

pacman::p_load(data.table, tidyverse, lmerTest, ggnewscale, mgcv, patchwork, broom.mixed)

base_path <- "C:/Users/PC/Documents/EvaluatingVDBA"

window_seconds <- 1 # redefine that here
 
dataset_variables <-read_csv(file.path(base_path, "Dataset_Variables.csv"), col_select = c(Name, Species, LogMass, Device, DeviceAttachment), show_col_types  = FALSE)

species_list <- list.dirs(file.path(base_path, "Data/AccelerometerData"), recursive = FALSE)

```

```{r plot_thme, include=FALSE}
my_theme <- function() {
  theme_minimal(base_size = 15, base_family = "serif") +
    theme(
      panel.border = element_rect(color = "black", linewidth = 1.5, fill = NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.title = element_text(size = 20),
      axis.text = element_text(size = 20) 
    )
}
fave_colours <- c(
  "firebrick3",
  "tan2",
  "goldenrod2", 
  "palegreen3",
  "aquamarine3",
  "cornflowerblue",
  "powderblue",
  "plum", 
  "rosybrown1",
  "lavenderblush2"
)

threshold_cols <- c(
  active   = "lightcoral", 
  all = "lightcyan3"
)

# plot functions
mean_max_plotting <- function(data){
  
  p1mean <- ggplot(data, aes(x = LogMass, y = log10(SpeciesMean), shape = threshold)) +
    geom_point(aes(colour = Name), size = 4) +
    xlab("Log mass (g)") + ylab("Log mean VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() + # reset the colour scale
    geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "bottom")
  
  return(p1mean)
}

quick_stats <- function(data){
  
  mean_model <- glmmTMB(log(meanVDBA) ~ LogMass * threshold + (1|ID), data = data)
  mean_res <- broom.mixed::tidy(mean_model, effects = "fixed")
  mean_results <- mean_res[grep("LogMass", mean_res$term), ]
  mean_results$sig <- ifelse(mean_results$p.value < 0.05, "sig", "not sig")
  mean_results$variable <- "mean"
  
  max_data <- data %>% dplyr::filter(threshold == "active")
  max_model <- glmmTMB(log(maxVDBA) ~ LogMass + (1|ID), data = max_data)
  max_res <- broom.mixed::tidy(max_model, effects = "fixed")
  max_results <- max_res[grep("LogMass", max_res$term), ]
  max_results$sig <- ifelse(max_results$p.value < 0.05, "sig", "not sig")
  max_results$variable <- "max"
  
  logmass_results <- rbind(mean_results, max_results)
  
  return(logmass_results)
}

quick_stats2 <- function(data){
  
  mean_model <- lm(log(SpeciesMean) ~ LogMass * threshold, data = data)
  mean_res <- broom.mixed::tidy(mean_model, effects = "fixed")
  mean_results <- mean_res[grep("LogMass", mean_res$term), ]
  mean_results$sig <- ifelse(mean_results$p.value < 0.05, "sig", "not sig")
  mean_results$variable <- "mean"
  
  max_data <- data %>% dplyr::filter(threshold == "active")
  max_model <- lm(log(SpeciesMeanMax) ~ LogMass, data = max_data)
  max_res <- broom.mixed::tidy(max_model, effects = "fixed")
  max_results <- max_res[grep("LogMass", max_res$term), ]
  max_results$sig <- ifelse(max_results$p.value < 0.05, "sig", "not sig")
  max_results$variable <- "max"
  
  logmass_results <- rbind(mean_results, max_results)
  
  return(logmass_results)
}

```


```{r}
# Prepare the data -------------------------------------------------------------

# this will include the 1 and 3 second windows 
summary_files1 <- list.files(file.path(base_path, "Data/AccelerometerData"), recursive = TRUE, pattern = "_summary\\.csv$", full.names = TRUE)

summary_data1 <- lapply(summary_files1, function(x){
  dat <- fread(x) %>% na.omit()
  dat$ID <- as.character(dat$ID)
  dat$Name <- basename(dirname(x))
  dat
})
summary_data1 <- rbindlist(summary_data1, fill = TRUE)

summary_data1 <- merge(summary_data1, dataset_variables, by = 'Name') %>%
  dplyr::filter(threshold != "inactive") %>%
  group_by(Name, threshold) %>%
  mutate(SpeciesMean = mean(meanVDBA), SpeciesMeanMax = mean(maxVDBA))
fwrite(summary_data1, file.path(base_path, "Output", paste0("All_summary_data.csv")))

```

## All species pooled
All species, with approximate/average masses. 
```{r, echo = FALSE}
# Plots ------------------------------------------------------------------------
plot1 <- mean_max_plotting(summary_data1)
plot1

stats1 <- quick_stats(summary_data1)
stats1

```

## Specific Masses
# Pooled
Maybe the species masses are too vauge... what if we only use the datasets where we know the exact mass?
Again, just a linear model with no random effect for ID.
```{r}
ind_species <- c("Annett_Kangaroo", "Clemente_Impala", "Clemente_Kudu", "Gaschk_Quoll", "Sparkes_Koala")

# Analysis ----------------------------------------------------------------
ind_summaries <- lapply(ind_species, function(x){
  weights <- fread(file.path(base_path, "Mass_of_individuals", paste0(x, ".csv"))) %>%
    mutate(ID = as.character(ID))
  summary <- fread(file.path(base_path, "Data/AccelerometerData", x, paste0(x, "_", window_seconds, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID))
  dat <- merge(summary, weights, by = "ID")
  dat
})
ind_data <- rbindlist(ind_summaries, fill = TRUE) %>%
  dplyr::filter(threshold != "inactive")

# Plots ------------------------------------------------------------------------------
 ggplot(ind_data, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
    geom_point(aes(colour = Dataset), size = 4) +
    xlab("Log mass (g)") + ylab("Log mean VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() + # reset the colour scale
    geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "bottom")

mean_model <- lm(log(meanVDBA) ~ LogMass * threshold, data = ind_data)
mean_res <- broom.mixed::tidy(mean_model, effects = "fixed")
mean_results <- mean_res[grep("LogMass", mean_res$term), ]
mean_results$sig <- ifelse(mean_results$p.value < 0.05, "sig", "not sig")
mean_results

```

# Even further separated
Where the masses of the exact individual are known. Per species plot and then all known individuals on the same graph.
```{r echo = FALSE}

ind_data <- ind_data %>%
  dplyr::filter(Dataset != "Clemente_Kudu")

ggplot(ind_data, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
    geom_point(aes(colour = Dataset), size = 3) +
    xlab("Log mass (g)") + ylab("Log mean VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() + # reset the colour scale
    geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "none") +
    facet_wrap(~Dataset, scales = "free")

```

## Understanding the distribution of activity
Does the distibution of high and low VDBA differ with size?

```{r}
# load in all the vdba data
all_files <- list.files(file.path(base_path, "Data/AccelerometerData"), recursive = TRUE, pattern = "_1_processed.csv", full.names = TRUE)
all_data <- lapply(all_files, function(x){
  dat <- fread(x)
  dat$Species <- basename(dirname(x))
  
  dat
})
all_data <- rbindlist(all_data)

# only include datasets with more than 10 hours (10 * 60 * 60)
selected_data <- all_data %>%
  group_by(Species) %>%
  dplyr::filter(n() > 5 * 60 * 60) %>% # 5 hours worth of data
  ungroup()

# now plot the distributions
ggplot(selected_data, aes(x = log10(seconds_VDBA), fill = Species)) + 
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 0.01) +
  scale_fill_manual(values = fave_colours) +
  scale_x_continuous(limits = c(-5, 1)) +
  labs(y = "Proportion of counts", x = "Log VDBA (g)") +
  my_theme() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 10)) +
  facet_wrap(~Species, scales = "free_y")

```

