---
title: "Results for VDBA Scaling Analysis"
author: "O.W."
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  out.width = "100%",
  fig.width = 12,
  fig.height = 6
)

pacman::p_load(data.table, tidyverse, lmerTest, ggnewscale, mgcv, patchwork)

base_path <- "C:/Users/PC/Documents/EvaluatingVDBA"

dataset_variables <- fread(file.path(base_path, "Dataset_Variables.csv"))
species_list <- list.dirs(file.path(base_path, "AccelerometerData"), recursive = FALSE)
maxxed_species <- c("Annett_Bettong", "Neis_Cow", "Pagano_Bear", "Sparkes_Koala", "Vehkaoja_Dog")

```

```{r plot_thme, include=FALSE}
my_theme <- function() {
  theme_minimal(base_size = 10, base_family = "serif") +
    theme(
      panel.border = element_rect(color = "black", linewidth = 1.5, fill = NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 12) 
    )
}
fave_colours <- c(
  "firebrick3", "tomato", "coral", "darkorange2", "sienna3", "tan2",
  "goldenrod2", "lemonchiffon2", "khaki2",
  "springgreen3", "darkseagreen4", "seagreen", "olivedrab4", "darkcyan", "aquamarine3",
  "skyblue4", "deepskyblue3", "cornflowerblue", "powderblue", "royalblue4", "slateblue2", "lightslateblue",
  "purple4", "mediumpurple3", "mediumorchid", "plum", "orchid3", "thistle",
  "deeppink3", "hotpink", "lightpink1", "rosybrown1",
  "mistyrose3", "lavenderblush2", "seashell2"
)

threshold_cols <- c(
  active   = "lightcoral", 
  inactive = "lightcyan3",
  all  = "ivory3"
)

```

# Methods

- 37 species datasets of raw tri-axial accelerometer data.
- Formatted to standard data structure
- Outliers manually removed through visual inspection and clipping
- Smoothed outliers with rolling mean (to remove unreasonable spikes)
- Generated dynamic VDBA by removing static acceleration (either as a rolling average of 10 seconds of VDBA for continuous data or as the average of the burst)
- Separated instances into active and inactive categories such that all instances with a rolling standard deviation. Everything above 0.05 VDBA was classed as "active"
- Summarised max and mean VDBA per individual.

## Individual Analysis
Where the masses of the exact individual are known. Per species plot and then all known individuals on the same graph.
```{r echo = FALSE}
species <- c("Annett_Kangaroo", "Clemente_Impala", "Gaschk_Quoll", "Sparkes_Koala", "Clemente_Kudu", "HarveyCaroll_Pangolin", "Vehkaoja_Dog")
species <- setdiff(species, maxxed_species)

# Analysis ----------------------------------------------------------------
all_summaries <- lapply(species, function(x){
  weights <- fread(file.path(base_path, "Mass_of_individuals", paste0(x, ".csv"))) %>%
    mutate(ID = as.character(ID))
  summary <- fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID))
  dat <- merge(summary, weights, by = "ID")
  dat
})

all_data <- rbindlist(all_summaries, fill = TRUE) %>%
  select(Dataset, ID, threshold, meanVDBA, minVDBA, maxVDBA, LogMass)

# Plots -------------------------------------------------------------------
p1 <- ggplot(all_data, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset, shape = threshold)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  
  facet_wrap(~Dataset, scales = "free") +
  my_theme()

p1.2 <- ggplot(all_data%>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
  geom_point(aes(colour = Dataset, shape = threshold)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  
  facet_wrap(~Dataset, scales = "free") +
  my_theme()

# and display
p1 + p1.2

# plot all the individuals togetehr
p2 <- ggplot(all_data, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset, shape = threshold)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()

p2.2 <- ggplot(all_data %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
  geom_point(aes(colour = Dataset, shape = threshold)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()

# and display these too
p2 + p2.2
```

Ran the stats on the mean and then the max. (Currently silenced as non-significant Mass effect)
```{r}
# Stats -------------------------------------------------------------------
# mean_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Dataset), data = all_data)
# summary(mean_model)
# 
# active_data <- all_data %>% dplyr::filter(threshold == "active")
# max_model <- lmer(log10(maxVDBA) ~ LogMass + (1 | Dataset), data = active_data)
# summary(max_model)

```

## Single system analysis
Where exact masses are not included (species averages) but all data is collected using the same accelerometer sampling protocol.Repeated for all Clemente datasets and for all Buchmann datasets. 
# V1: Axivity (Clemente)
```{r echo = FALSE}
axivity_species <- dataset_variables$Name[dataset_variables$Device == "Axivity" & dataset_variables$Type == "Mam"]
axivity_species <- setdiff(axivity_species, maxxed_species)

# Analysis ----------------------------------------------------------------
axivity_summaries <- lapply(axivity_species, function(x){
 fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID),
           LogMass = dataset_variables$LogMass[dataset_variables$Name == x]
           )
})
axivity_summaries <- rbindlist(axivity_summaries, fill = TRUE) %>%
  select(Dataset, ID, threshold, meanVDBA, minVDBA, maxVDBA, LogMass)

# Plots -------------------------------------------------------------------
p3 <- ggplot(axivity_summaries, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()

p3.2 <- ggplot(axivity_summaries %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()

# display
p3 + p3.2
```

Run the stats, mean then max. (Silenced because non-significant again)
```{r}
# Stats -------------------------------------------------------------------
# axivity_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Dataset), data = axivity_summaries)
# summary(axivity_model)
# 
# active_data <- axivity_summaries %>% dplyr::filter(threshold == "active")
# max_model <- lmer(log10(maxVDBA) ~ LogMass + (1 | Dataset), data = active_data)
# summary(max_model)
```
# V2: Buchmann
```{r echo = FALSE}
buchmann_species <- dataset_variables$Name[str_detect(dataset_variables$Name, "Buchmann") %in% TRUE]

# Analysis ----------------------------------------------------------------
buchmann_summaries <- lapply(buchmann_species, function(x){
 fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID),
           LogMass = dataset_variables$LogMass[dataset_variables$Name == x]
           )
})
buchmann_summaries <- rbindlist(buchmann_summaries, fill = TRUE) %>%
  select(Dataset, ID, threshold, meanVDBA, minVDBA, maxVDBA, LogMass)

# Plots -------------------------------------------------------------------
p3.3 <- ggplot(buchmann_summaries, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()

p3.4 <- ggplot(buchmann_summaries %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()


p3.3+ p3.4
```

Stats displayed here because the maxVDBA was significant for LogMass.
```{r}
# Stats -------------------------------------------------------------------
buchmann_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Dataset), data = buchmann_summaries)
summary(buchmann_model)

buchmann_active_data <- buchmann_summaries %>% dplyr::filter(threshold == "active")
buchmann_max_model <- lmer(log10(maxVDBA) ~ LogMass + (1 | Dataset), data = buchmann_active_data)
summary(buchmann_max_model)
```

## All species pooled
All species, with approximate/average masses. 
```{r, echo = FALSE}
all_summary_files <- list.files(file.path(base_path, "AccelerometerData"), recursive = TRUE, pattern = "*_summary\\.csv$", full.names = TRUE)

vdba_data <- lapply(all_summary_files, function(x){
  dat <- fread(x) %>% na.omit()
  dat$ID <- as.character(dat$ID)
  dat$Name <- basename(dirname(x))
  dat
})
vdba_data <- rbindlist(vdba_data, fill = TRUE)

vdba_stuff1 <- merge(vdba_data, dataset_variables, by = 'Name')
fwrite(vdba_stuff1, file.path(base_path, "Output", "All_summary_data.csv"))

```

Only datasets with more than 24 hours of data (more likely to capture representative data).
```{r}
# select only the big datasets
all_volumes <- lapply(species_list, function(x){
  vol <- fread(file.path(x, paste0(basename(x), "_volume.csv")))
  vol$Species <- basename(x)
  vol
})
all_volumes <- rbindlist(all_volumes)
big_volumes <- all_volumes[all_volumes$volume > 24, ]
big_volumes_data <- vdba_stuff1 %>% dplyr::filter(!Name %in% maxxed_species)

vdba_stuff1 <- big_volumes_data

vdba_stuff1 <- vdba_stuff1 %>%
  select(ID, threshold, meanVDBA, minVDBA, maxVDBA, Name, LogMass, Frequency, DeviceAttachment) %>%
  dplyr::filter(!Name == "Chakravarty_Meerkat")

# summarise by dataset
vdba_stuff1_summarised <- vdba_stuff1 %>% group_by(Name, threshold, LogMass) %>%
   summarise(meanVDBA = mean(meanVDBA),
             maxVDBA = mean(maxVDBA),
             minVDBA = mean(minVDBA)
             )

# Plots ------------------------------------------------------------------------
p4 <- ggplot(vdba_stuff1_summarised, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme() +
  theme(legend.position = "none")

p5 <- ggplot(vdba_stuff1_summarised %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme() +
  theme(legend.position = "none")


p4 + p5

# stats ------------------------------------------------------------------------
all_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + Frequency + (1 | Name) + (1|DeviceAttachment), data = vdba_stuff1)
summary(all_model)

```

And then only the files where the system didn't max out.
```{r echo = FALSE}
# non-maxed out files
range_species <- setdiff(basename(species_list), maxxed_species)

vdba_stuff2_summarised <- vdba_stuff1_summarised %>% dplyr::filter(Name %in% range_species)

p7 <- ggplot(vdba_stuff2_summarised, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme() + 
  theme(legend.position = "none")

p8 <- ggplot(vdba_stuff2_summarised, aes(x = LogMass, y = log10(maxVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()  + 
  theme(legend.position = "none")

p7 + p8

all_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Name), data = vdba_stuff2_summarised)
summary(all_model)
```

## Find the active minutes per day
Are large animals more or less active generally?
```{r}

all_minutes_files <- list.files(file.path(base_path, "AccelerometerData"), recursive = TRUE, pattern = "*_active_minutes\\.csv$", full.names = TRUE)

minute_data <- lapply(all_minutes_files, function(x){
  dat <- fread(x) %>% na.omit()
  dat$Name <- basename(dirname(x))
  dat
})
minute_data <- rbindlist(minute_data, fill = TRUE)

dat <- merge(vdba_stuff1, minute_data, by = c("ID", "Name"))

# plot this
ggplot(dat, aes(x = LogMass, y = prop_active, colour = Name, group = 1)) + geom_point() + geom_smooth(method = "lm")

```
