---
title: "Results for VDBA Scaling Analysis"
author: "O.W."
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.width = 13,
  fig.height = 6
)

pacman::p_load(data.table, tidyverse, lmerTest, ggnewscale, mgcv, patchwork, broom.mixed)

base_path <- "C:/Users/PC/Documents/EvaluatingVDBA"
 
dataset_variables <- fread(file.path(base_path, "Dataset_Variables.csv"))
species_list <- list.dirs(file.path(base_path, "AccelerometerData"), recursive = FALSE)

```

```{r plot_thme, include=FALSE}
my_theme <- function() {
  theme_minimal(base_size = 10, base_family = "serif") +
    theme(
      panel.border = element_rect(color = "black", linewidth = 1.5, fill = NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.title = element_text(size = 20),
      axis.text = element_text(size = 15) 
    )
}
fave_colours <- c(
  "firebrick3", "tomato", "coral", "darkorange2", "sienna3", "tan2",
  "goldenrod2", "lemonchiffon2", "khaki2",
  "springgreen3", "darkseagreen4", "seagreen", "olivedrab4", "darkcyan", "aquamarine3",
  "skyblue4", "deepskyblue3", "cornflowerblue", "powderblue", "royalblue4", "slateblue2", "lightslateblue",
  "purple4", "mediumpurple3", "mediumorchid", "plum", "orchid3", "thistle",
  "deeppink3", "hotpink", "lightpink1", "rosybrown1",
  "mistyrose3", "lavenderblush2", "seashell2"
)

threshold_cols <- c(
  active   = "lightcoral", 
  inactive = "lightcyan3",
  all  = "ivory3"
)

# plot functions
mean_max_plotting <- function(data){
  
  p1mean <- ggplot(data, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
    geom_point(aes(colour = Name)) +
    xlab("Log mass (g)") + ylab("Log mean VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() + # reset the colour scale
    geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "none")
  
  p1max <- ggplot(data %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
    geom_point(aes(colour = Name)) +
    xlab("Log mass (g)") + ylab("Log max VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() +
    geom_smooth(method = "lm", colour = "lightcoral", se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "none")
  
  plot <- (p1mean + p1max) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10),
      legend.text  = element_text(size = 10),
      legend.key.size = unit(0.4, "lines"),
      legend.spacing.x = unit(0.3, "cm"),
      legend.spacing.y = unit(0.2, "cm")
    )
  
  return(plot)
}

quick_stats <- function(data){
  
  mean_model <- lmer(log(meanVDBA) ~ LogMass * threshold + (1|ID), data = data)
  mean_res <- broom.mixed::tidy(mean_model, effects = "fixed")
  mean_results <- mean_res[grep("LogMass", mean_res$term), ]
  mean_results$sig <- ifelse(mean_results$p.value < 0.05, "sig", "not sig")
  mean_results$variable <- "mean"
  
  max_data <- data %>% dplyr::filter(threshold == "active")
  max_model <- lmer(log(maxVDBA) ~ LogMass + (1|ID), data = max_data)
  max_res <- broom.mixed::tidy(max_model, effects = "fixed")
  max_results <- max_res[grep("LogMass", max_res$term), ]
  max_results$sig <- ifelse(max_results$p.value < 0.05, "sig", "not sig")
  max_results$variable <- "max"
  
  logmass_results <- rbind(mean_results, max_results)
  
  return(logmass_results)
}

quick_stats2 <- function(data){
  
  mean_model <- lm(log(meanVDBA) ~ LogMass * threshold, data = data)
  mean_res <- broom.mixed::tidy(mean_model, effects = "fixed")
  mean_results <- mean_res[grep("LogMass", mean_res$term), ]
  mean_results$sig <- ifelse(mean_results$p.value < 0.05, "sig", "not sig")
  mean_results$variable <- "mean"
  
  max_data <- data %>% dplyr::filter(threshold == "active")
  max_model <- lm(log(maxVDBA) ~ LogMass, data = max_data)
  max_res <- broom.mixed::tidy(max_model, effects = "fixed")
  max_results <- max_res[grep("LogMass", max_res$term), ]
  max_results$sig <- ifelse(max_results$p.value < 0.05, "sig", "not sig")
  max_results$variable <- "max"
  
  logmass_results <- rbind(mean_results, max_results)
  
  return(logmass_results)
}

```

```{r}
# Prepare the data -------------------------------------------------------------
all_summary_files <- list.files(file.path(base_path, "AccelerometerData"), recursive = TRUE, pattern = "*_summary\\.csv$", full.names = TRUE)

summary_data <- lapply(all_summary_files, function(x){
  dat <- fread(x) %>% na.omit()
  dat$ID <- as.character(dat$ID)
  dat$Name <- basename(dirname(x))
  dat
})
summary_data <- rbindlist(summary_data, fill = TRUE)

summary_data <- merge(summary_data, dataset_variables, by = 'Name') %>%
  dplyr::filter(!Name == "Chakravarty_Meerkat") %>%
  mutate(mean_MSVDBA = meanVDBA / 10^LogMass,
         max_MSVDBA = maxVDBA / 10^LogMass)
fwrite(summary_data, file.path(base_path, "Output", "All_summary_data.csv"))

```


## All species pooled
All species, with approximate/average masses. 
```{r, echo = FALSE}
# Plots ------------------------------------------------------------------------
plot1 <- mean_max_plotting(summary_data)
plot1

stats1 <- quick_stats(summary_data)
stats1
```

## Only Datasets With Long Data
Maybe some of the small datasets are skewing it... Exclude datasets that had less than 24 hrs worth of data.
```{r}
# select only the big datasets
all_volumes <- lapply(species_list, function(x){
  vol <- fread(file.path(x, paste0(basename(x), "_volume.csv")))
  vol$Species <- basename(x)
  vol
})
all_volumes <- rbindlist(all_volumes)
long_volumes <- all_volumes[all_volumes$volume > 12, ]
long_species <- unique(long_volumes$Species)
long_volumes_data <- summary_data %>% dplyr::filter(Name %in% long_species)

plot2 <- mean_max_plotting(long_volumes_data)
plot2

stats2 <- quick_stats(long_volumes_data)
stats2

```

## Collar Data Only
And then only the sepcies that were on a collar.
```{r}
collar_data <- summary_data %>% dplyr::filter(DeviceAttachment == "Collar")

plot3 <- mean_max_plotting(collar_data)
plot3

stats3 <- quick_stats(collar_data)
stats3

```

## Single system analysis
What is the differences in sampling protocol are throwing it out? Collected only data collected in the same way by the same lab.
# V1: Clemente
There were insufficint individuals, so removed ID as an effect and just had a linear model.
```{r echo = FALSE}
Clemente_data <- summary_data %>% dplyr::filter(Device == "Axivity")

plot4 <- mean_max_plotting(Clemente_data)
plot4

stats4 <- quick_stats2(Clemente_data)
stats4
```

# V2: Buchmann
Similarly ID was not a random effect here as too few individuals.
```{r echo = FALSE}
buchmann_species <- dataset_variables$Name[str_detect(dataset_variables$Name, "Buchmann") %in% TRUE]
Buchmann_data <- summary_data %>% dplyr::filter(Name %in% buchmann_species)

plot5 <- mean_max_plotting(Buchmann_data)
plot5

stats5 <- quick_stats2(Buchmann_data)
stats5
```

## Specific Masses
# Pooled
Maybe the species masses are too vauge... what if we only use the datasets where we know the exact mass?
Again, just a linear model with no random effect for ID.
```{r}
ind_species <- c("Annett_Kangaroo", "Clemente_Impala", "Gaschk_Quoll", "HarveyCaroll_Pangolin",  "Clemente_Kudu")
# "Annett_Bettong", "Annett_Possum", "Annett_Glider", 

# Analysis ----------------------------------------------------------------
ind_summaries <- lapply(ind_species, function(x){
  weights <- fread(file.path(base_path, "Mass_of_individuals", paste0(x, ".csv"))) %>%
    mutate(ID = as.character(ID))
  summary <- fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID))
  dat <- merge(summary, weights, by = "ID")
  dat
})

ind_data <- rbindlist(ind_summaries, fill = TRUE)

# Plots ------------------------------------------------------------------------------
 p6mean <- ggplot(ind_data, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
    geom_point(aes(colour = Dataset)) +
    xlab("Log mass (g)") + ylab("Log mean VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() + # reset the colour scale
    geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "none")
  
  p6max <- ggplot(ind_data %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
    geom_point(aes(colour = Dataset)) +
    xlab("Log mass (g)") + ylab("Log max VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() +
    geom_smooth(method = "lm", colour = "lightcoral", se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "none")
  
  plot6 <- (p6mean + p6max) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10),
      legend.text  = element_text(size = 10),
      legend.key.size = unit(0.4, "lines"),
      legend.spacing.x = unit(0.3, "cm"),
      legend.spacing.y = unit(0.2, "cm")
    )

plot6

stats6 <- quick_stats2(ind_data)
stats6

```

# Even further separated
Where the masses of the exact individual are known. Per species plot and then all known individuals on the same graph.
```{r echo = FALSE}

# Plots -------------------------------------------------------------------
  p7mean <- ggplot(ind_data, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
    geom_point(aes(colour = Dataset)) +
    xlab("Log mass (g)") + ylab("Log mean VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() + # reset the colour scale
    geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "none") +
    facet_wrap(~Dataset, scales = "free")
  
  p7max <- ggplot(ind_data %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
    geom_point(aes(colour = Dataset)) +
    xlab("Log mass (g)") + ylab("Log max VDBA (g)") +
    scale_color_manual(values = fave_colours) +
    ggnewscale::new_scale_colour() +
    geom_smooth(method = "lm", colour = "lightcoral", se = FALSE, linewidth = 1.2) +
    scale_color_manual(values = threshold_cols) +
    my_theme() +
    theme(legend.position = "none")+
    facet_wrap(~Dataset, scales = "free")
  
  plot7 <- (p7mean + p7max) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10),
      legend.text  = element_text(size = 10),
      legend.key.size = unit(0.4, "lines"),
      legend.spacing.x = unit(0.3, "cm"),
      legend.spacing.y = unit(0.2, "cm")
    )

plot7

```

## Understanding the animal activity
# Find the active minutes per day
Is there a mass-mediated effect on activity patterns generally
```{r}
# find the minutes per day that the animals are active
all_minutes_files <- list.files(file.path(base_path, "AccelerometerData"), recursive = TRUE, pattern = "*_active_minutes\\.csv$", full.names = TRUE)
minute_data <- lapply(all_minutes_files, function(x){
  dat <- fread(x) %>% na.omit()
  dat$Name <- basename(dirname(x))
  dat
})
minute_data <- rbindlist(minute_data, fill = TRUE)
minute_data <- minute_data %>% dplyr::filter(Name %in% long_species)
merged_data <- merge(summary_data, minute_data, by = c("ID", "Name"))

# plot this
merged_data$prop_active <- 1 - merged_data$prop_rest
p8total <- ggplot(merged_data, aes(x = LogMass, y = prop_active, colour = Name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Log Mass (g)") +
  ylab("Proportion of timeseries active") +
  scale_color_manual(values = fave_colours) +
  my_theme() +
  theme(legend.position = "bottom")

# average it across the individuals
plot_data <- merged_data %>%
  group_by(Name, LogMass) %>%
  summarise(mean_high = mean(prop_high),
            mean_med = mean(prop_med),
            mean_low = mean(prop_low),
            mean_rest = mean(prop_rest))

# rotate it to allow all the activity levels
plot_data <- plot_data %>%
  pivot_longer(cols = c(mean_high, mean_med, mean_low, mean_rest),
               names_to = "activity",
               values_to = "prop")
plot_data$activity <- factor(plot_data$activity, levels = c("mean_high", "mean_med", "mean_low", "mean_rest"))

p8props <- ggplot(plot_data,  aes(x = LogMass, y = prop, colour = activity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Log Mass (g)") +
  ylab("Proportion of timeseries per activity level") +
  scale_colour_manual(values = c("mean_high" = "lightcoral",
                               "mean_med" = "tan2",
                               "mean_low" = "lightcyan3",
                               "mean_rest" = "ivory3")) +
  my_theme() +
  theme(legend.position = "bottom")

plot8 <- (p8total + p8props) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10),
      legend.text  = element_text(size = 10),
      legend.key.size = unit(0.4, "lines"),
      legend.spacing.x = unit(0.3, "cm"),
      legend.spacing.y = unit(0.2, "cm")
    )

plot8

```
