---
title: "Results for VDBA Scaling Analysis"
author: "O.W."
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(data.table, tidyverse, lmerTest, ggnewscale, mgcv)

base_path <- "C:/Users/PC/Documents/EvaluatingVDBA"

dataset_variables <- fread(file.path(base_path, "Dataset_Variables.csv"))
species_list <- list.dirs(file.path(base_path, "AccelerometerData"), recursive = FALSE)
maxxed_species <- c("Annett_Bettong", "Neis_Cow", "Pagano_Bear", "Sparkes_Koala", "Vehkaoja_Dog")


```

```{r plot_thme, include=FALSE}
my_theme <- function() {
  theme_minimal(base_size = 10, base_family = "serif") +
    theme(
      panel.border = element_rect(color = "black", linewidth = 1.5, fill = NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 12) 
    )
}
fave_colours <- c(
  "firebrick3", "tomato", "coral", "darkorange2", "sienna3", "tan2",
  "goldenrod2", "lemonchiffon2", "khaki2",
  "springgreen3", "darkseagreen4", "seagreen", "olivedrab4", "darkcyan", "aquamarine3",
  "skyblue4", "deepskyblue3", "cornflowerblue", "powderblue", "royalblue4", "slateblue2", "lightslateblue",
  "purple4", "mediumpurple3", "mediumorchid", "plum", "orchid3", "thistle",
  "deeppink3", "hotpink", "lightpink1", "rosybrown1",
  "mistyrose3", "lavenderblush2", "seashell2"
)

threshold_cols <- c(
  active   = "lightcoral", 
  inactive = "lightcyan3",
  all  = "ivory3"
)

```

# Methods

- 37 species datasets of raw tri-axial accelerometer data.
- Formatted to standard data structure
- Smoothed outliers with a butterworth filter (cutoff_freq, 0.5 Hz, to remove only sudden spikes)
- Generated dynamic VDBA by removing static acceleration (either as a rolling average of 10 seconds of VDBA for continuous data or as the average of the burst)
- Separated instances into active and inactive categories such that all instances with a rolling standard deviation (measure of flatness in the rolling window) in the botton 25% were classed as inactive.
- Summarised max and mean VDBA per individual.

## Individual Analysis
Where the masses of the exact individual are known. Per species plot and then all known individuals on the same graph.
```{r echo = FALSE}
species <- c("Annett_Kangaroo", "Clemente_Impala", "Gaschk_Quoll", "Sparkes_Koala", "Clemente_Kudu", "HarveyCaroll_Pangolin") # "Vehkaoja_Dog"

# Analysis ----------------------------------------------------------------
all_summaries <- lapply(species, function(x){
  weights <- fread(file.path(base_path, "Mass_of_individuals", paste0(x, ".csv"))) %>%
    mutate(ID = as.character(ID))
  summary <- fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID))
  dat <- merge(summary, weights, by = "ID")
  dat
})

all_data <- rbindlist(all_summaries, fill = TRUE) %>%
  select(Dataset, ID, threshold, meanVDBA, minVDBA, maxVDBA, LogMass)

# Plots -------------------------------------------------------------------
p1 <- ggplot(all_data, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset, shape = threshold)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  
  facet_wrap(~Dataset, scales = "free") +
  my_theme()
p1

p1.2 <- ggplot(all_data, aes(x = LogMass, y = log10(maxVDBA))) +
  geom_point(aes(colour = Dataset, shape = threshold)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  
  facet_wrap(~Dataset, scales = "free") +
  my_theme()
p1.2

# plot all the individuals togetehr
p2 <- ggplot(all_data, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset, shape = threshold)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p2

p2.2 <- ggplot(all_data %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
  geom_point(aes(colour = Dataset, shape = threshold)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p2.2

# Stats -------------------------------------------------------------------
mean_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Dataset), data = all_data)
summary(mean_model)

active_data <- all_data %>% filter(threshold == "active")
max_model <- lmer(log10(maxVDBA) ~ LogMass + (1 | Dataset), data = active_data)
summary(max_model)

```

## Single system analysis
Where exact masses are not included (species averages) but all data is collected using the same accelerometer sampling protocol.Repeated for all Clemente datasets and for all Buchmann datasets. 
# V1: Axivity (Clemente)
```{r echo = FALSE}
axivity_species <- dataset_variables$Name[dataset_variables$Device == "Axivity" & dataset_variables$Type == "Mam"]
axivity_species <- axivity_species[-6]

# Analysis ----------------------------------------------------------------
axivity_summaries <- lapply(axivity_species, function(x){
 fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID),
           LogMass = dataset_variables$LogMass[dataset_variables$Name == x]
           )
})
axivity_summaries <- rbindlist(axivity_summaries, fill = TRUE) %>%
  select(Dataset, ID, threshold, meanVDBA, minVDBA, maxVDBA, LogMass)

# Plots -------------------------------------------------------------------
p3 <- ggplot(axivity_summaries, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p3

p3.2 <- ggplot(axivity_summaries %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(maxVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p3.2

# Stats -------------------------------------------------------------------
axivity_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Dataset), data = axivity_summaries)
summary(axivity_model)

```
# V2: Buchmann
```{r echo = FALSE}
buchmann_species <- dataset_variables$Name[str_detect(dataset_variables$Name, "Buchmann") %in% TRUE]

# Analysis ----------------------------------------------------------------
buchmann_summaries <- lapply(buchmann_species, function(x){
 fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID),
           LogMass = dataset_variables$LogMass[dataset_variables$Name == x]
           )
})
buchmann_summaries <- rbindlist(buchmann_summaries, fill = TRUE) %>%
  select(Dataset, ID, threshold, meanVDBA, minVDBA, maxVDBA, LogMass)

# Plots -------------------------------------------------------------------
p3.3 <- ggplot(buchmann_summaries, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p3.3

p3.4 <- ggplot(buchmann_summaries %>% dplyr::filter(threshold == "active"), aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p3.4

# Stats -------------------------------------------------------------------
buchmann_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Dataset), data = buchmann_summaries)
summary(buchmann_model)

# buchmann_active_data <- buchmann_summaries %>% dplyr::filter(threshold == "active")
# buchmann_max_model <- lmer(log10(maxVDBA) ~ LogMass + (1 | Dataset), data = buchmann_active_data)
# summary(buchmann_max_model)

```

## All species pooled
All species, with approximate/average masses. 
```{r, echo = FALSE}
all_summary_files <- list.files(file.path(base_path, "AccelerometerData"), recursive = TRUE, pattern = "*_summary\\.csv$", full.names = TRUE)

vdba_data <- lapply(all_summary_files, function(x){
  dat <- fread(x) %>% na.omit()
  dat$ID <- as.character(dat$ID)
  dat$Name <- basename(dirname(x))
  dat
})
vdba_data <- rbindlist(vdba_data, fill = TRUE)

vdba_stuff1 <- merge(vdba_data, dataset_variables, by = 'Name')

vdba_stuff1 <- vdba_stuff1 %>%
  select(ID, threshold, meanVDBA, minVDBA, maxVDBA, Name, LogMass, Frequency, DeviceAttachment) %>%
  dplyr::filter(!Name == "Chakravarty_Meerkat")

fwrite(vdba_stuff1, file.path(base_path, "Output", "All_summary_data.csv"))

# summarise by dataset
vdba_stuff1_summarised <- vdba_stuff1 %>% group_by(Name, threshold, LogMass) %>%
   summarise(meanVDBA = mean(meanVDBA),
             maxVDBA = mean(maxVDBA),
             minVDBA = mean(minVDBA)
             )

# Plots ------------------------------------------------------------------------
p4 <- ggplot(vdba_stuff1_summarised, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme() 
p4 

p5 <- ggplot(vdba_stuff1_summarised, aes(x = LogMass, y = log10(maxVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p5

# stats ------------------------------------------------------------------------
all_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + Frequency + (1 | Name) + (1|DeviceAttachment), data = vdba_stuff1)
summary(all_model)

```

And then only the files where the system didn't max out.
```{r echo = FALSE}
# non-maxed out files
range_species <- setdiff(basename(species_list), maxxed_species)

vdba_stuff2_summarised <- vdba_stuff1_summarised %>% dplyr::filter(Name %in% range_species)

p7 <- ggplot(vdba_stuff2_summarised, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme() 
p7 

p8 <- ggplot(vdba_stuff2_summarised, aes(x = LogMass, y = log10(maxVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p8
```



## Curvi-linear trend
Chris says: "I wonder if you picked only the highest VDBA for each body mass, and used a GAMS model to predict the pattern, would it be hump shaped? with a peak around 50kg? A curvilinear trend might also explain why sometimes we see a positive slope (e.g. with quolls, small size) and a negative slope with kangaroos (large body size)"

```{r echo = FALSE}

# summarise by dataset
vdba_stuff_max <- vdba_stuff1 %>% group_by(Name, threshold, LogMass) %>%
   summarise(maxVDBA = max(maxVDBA)) %>%
  dplyr::filter(threshold == "active")

# Plots ------------------------------------------------------------------------
p6 <- ggplot(vdba_stuff_max, aes(x = LogMass, y = log10(maxVDBA))) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() +  # reset the colour scale
  geom_smooth(method = "gam", colour = "ivory3", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p6


# stats ------------------------------------------------------------------------
gam_model <- gam(log10(maxVDBA) ~ s(LogMass), data = vdba_stuff_max)
summary(gam_model)

```
