---
title: "Results for VDBA Scaling Analysis"
author: "O.W."
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(data.table, tidyverse, lmerTest, ggnewscale)

base_path <- "C:/Users/PC/Documents/EvaluatingVDBA"

dataset_variables <- fread(file.path(base_path, "Dataset_Variables.csv"))
```

```{r plot_thme, include=FALSE}
my_theme <- function() {
  theme_minimal(base_size = 10, base_family = "serif") +
    theme(
      panel.border = element_rect(color = "black", linewidth = 1.5, fill = NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 12) 
    )
}
fave_colours <- c(
  "firebrick3", "tomato", "coral", "lightsalmon", "sienna3", "tan2",
  "goldenrod2", "lemonchiffon2", "khaki2",
  "darkseagreen4", "seagreen", "aquamarine3", "darkcyan",
  "skyblue4", "deepskyblue3", "cornflowerblue", "powderblue",
  "royalblue4", "slateblue2", "lightslateblue",
  "purple4", "mediumpurple3", "mediumorchid", "plum", "orchid3", "thistle",
  "deeppink3", "lightpink1", "mistyrose3", "lavenderblush2", "seashell2"
)

threshold_cols <- c(
  active   = "lightcoral", 
  inactive = "lightcyan3",
  all  = "ivory3"
)

```

## PART 1: Individual Analysis
Where the masses of the exact individual are known. Per species plot and then all known individuals on the same graph.
```{r 1, echo = FALSE}
species <- c("Annett_Kangaroo", "Clemente_Impala", "Gaschk_Quoll", "Sparkes_Koala", "Clemente_Kudu")

# Analysis ----------------------------------------------------------------
all_summaries <- lapply(species, function(x){
  weights <- fread(file.path(base_path, "AccelerometerData", x, "Mass_of_Individuals.csv")) %>%
    mutate(ID = as.character(ID))
  summary <- fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID))
  dat <- merge(summary, weights, by = "ID")
  dat
})

all_data <- rbindlist(all_summaries, fill = TRUE) %>%
  select(Dataset, ID, threshold, meanVDBA, minVDBA, maxVDBA, LogMass)

# Plots -------------------------------------------------------------------
p1 <- ggplot(all_data, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  
  facet_wrap(~Dataset, scales = "free") +
  my_theme()
p1

# plot all the individuals togetehr
p2 <- ggplot(all_data, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p2

# Stats -------------------------------------------------------------------
model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Dataset), data = all_data)
summary(model)

```


## PART 2: Axivity analysis
Where exact masses are not included (species averages) but all data is collected using the same accelerometer (Axivity AX3 and AX6) sampling protocol.

```{r 2, echo = FALSE}
axivity_species <- dataset_variables$Name[dataset_variables$Device == "Axivity"]

# Analysis ----------------------------------------------------------------
axivity_summaries <- lapply(axivity_species, function(x){
 fread(file.path(base_path, "AccelerometerData", x, paste0(x, "_summary.csv"))) %>%
    mutate(Dataset = x,
           ID = as.character(ID),
           LogMass = dataset_variables$LogMass[dataset_variables$Name == x]
           )
})
axivity_summaries <- rbindlist(axivity_summaries, fill = TRUE) %>%
  select(Dataset, ID, threshold, meanVDBA, minVDBA, maxVDBA, LogMass)

# Plots -------------------------------------------------------------------
p3 <- ggplot(axivity_summaries, aes(x = LogMass, y = log10(meanVDBA))) +
  geom_point(aes(colour = Dataset)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p3

# Stats -------------------------------------------------------------------
axivity_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Dataset), data = axivity_summaries)
summary(axivity_model)

```

## PART 3: All Species
All species, with approximate/average masses. 
```{r, echo = FALSE}
all_summary_files <- list.files(file.path(base_path, "AccelerometerData"), recursive = TRUE, pattern = "*_summary\\.csv$", full.names = TRUE)

vdba_data <- lapply(all_summary_files, function(x){
  dat <- fread(x) %>% na.omit()
  dat$ID <- as.character(dat$ID)
  dat$Name <- basename(dirname(x))
  dat
})
vdba_data <- rbindlist(vdba_data, fill = TRUE)

vdba_data <- vdba_data %>%
  select(ID, threshold, meanVDBA, minVDBA, maxVDBA, Name) %>%
  dplyr::filter(!Name == "Chakravarty_Meerkat")

vdba_stuff1 <- merge(vdba_data, dataset_variables, by = 'Name')

# summarise by dataset
vdba_stuff1_summarised <- vdba_stuff1 %>% group_by(Name, threshold, LogMass) %>%
   summarise(meanVDBA = mean(meanVDBA),
             maxVDBA = mean(maxVDBA),
             minVDBA = mean(minVDBA)
             )

# Plots ------------------------------------------------------------------------
p4 <- ggplot(vdba_stuff1, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme() 
p4 

p5 <- ggplot(vdba_stuff1_summarised, aes(x = LogMass, y = log10(meanVDBA), shape = threshold)) +
  geom_point(aes(colour = Name)) +
  scale_color_manual(values = fave_colours) +
  ggnewscale::new_scale_colour() + # reset the colour scale
  geom_smooth(method = "lm", aes(colour = threshold), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = threshold_cols) +
  my_theme()
p5

# stats ------------------------------------------------------------------------
all_model <- lmer(log10(meanVDBA) ~ LogMass * threshold + (1 | Name) + (1|DeviceAttachment), data = vdba_stuff1)
summary(all_model)

```





